---
title: "Testing for Mortality Differences in Three Strains of Insect"
author: "Dan Hartman"
date: "2025-03-28"
output: rmarkdown::github_document
output_file: "README"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include=FALSE, echo=FALSE}
library(tidyverse)
library(emmeans)
setwd("C:/Users/hartm/OneDrive/Documents/bioassay_analysis_workshop/Meeting_2")
```

## Objectives
1.  Understand the basics of plotting with ggplot()
2.  Use an example dataset to fit a binomial regression model.
3.  Use the model to get some information back about our dataset.

Take a minute to familiarize yourself with the script for todays meetup. It's
long, but most of that is text to explain what is going on. Going forward,
there will be more code and fewer comments. The remainder of this  document is
redundant with the comments in the R script provided, in case you don't want to
toggle back and forth.

Also take a minute to look at the example dataset (three.strain.mortality.csv).
This represents three strains of insect, in a situation where you are testing
for differences in mortality. This would be with a single insecticide, and with
a uniform dose. I'm imagining bottle assays here (thus the bottle column).

## The Workflow
These data were simulated, meaning that I chose the mortality rates for each
strain, and used some R code to generate bioassay results according to those
rates. So, there is naturally a little bit of noise. This dataset is what you
will read in to R.

The next step is to fit a model to those simulated data. Think of this as
drawing a line through those data points, just as you would with a regular
linear regression. This model is better specified to data that we have,
which are binomial rather than normal or Gaussian.

Finally, we can get some inference from the model output concerning differences
in mortality rates between the strains. Using the parameters from the fitted
model, we can back-estimate the mortality rates and compare them to the ones
we specified when we simulated the data. In an experimental setting you won't
have the luxury of that prior information, of course. These comparisons will
serve as a sanity check, and to show that we did the right things in our code.

## Excercise

### Get organized.  Future-you will thank you.
Don't forget to make a project folder. Put the example data set into this
folder, then the R script, then open the R script in RStudio. Take a minute to
set your working directory to this folder. You can do that in the bottom-right
window (Files tab), or like this (your path will differ):

```{r, eval=FALSE}
setwd("C:/Users/elvispresley/Documents/stats_project")
```

### Data Exploration
Now you are ready to read in the data and do some checks:
```{r}
data <- read_csv("three.strain.mortality.csv")
str(data) #This time we have double integers as well as character columns
head(data) #Look at the top few rows of the table
```

Here is a data cleaning step:
```{r}
unique(data$strain) #get the unique values for each strain
unique(data$treatment) #get unique bottle types
```
If we have any typos in the sheet, we will see them here and correct them before
going any further.

What are we missing?  Mortality rates of course!  Lets calculate those. mutate()
can add a column calculated from other columns.
```{r}
data<-mutate(data, mort.rate=dead/total.mosquitoes) #we are overwriting "data"
head(data)
```


How to look at only negative controls? This is called subsetting, or filtering.
Great data exploration tool. There are ways with base R, but tidyverse makes
this a lot easier for us.
```{r}
filter(data, treatment=="negative.control")
```

Notice the double "=".  this is necessary for logical equations. Great, low
mortality in control bottles.  Not bad!

An equivalent way of filtering for negative control bottles:
```{r}
data %>% filter(treatment=="negative.control")
```
The %>% takes the data and passes it through one or several steps of manipulation.

We can also have multiple filtering criteria:
```{r}
data %>% filter(treatment=="negative.control", strain=="dos")
```

Let's make a quick plot of the negative control bottles using base R:
```{r}
neg.bottles<-data %>% filter(treatment=="negative.control")
plot(y=neg.bottles$mort.rate, x=as.factor(neg.bottles$strain), type="p")
```

It works, but I don't like it.  Let's get some better plotting going with ggplot.
This graphics package loads automatically with tidyverse, so we are good to go.
```{r}
ggplot(data=data) +
  geom_point(aes(x=strain, y=mort.rate)) #aes() is for "aesthetics"
```

That's cool, but how do we distinguish between treated and negative control bottles?
```{r}
ggplot(data=data) +
  geom_point(aes(x=strain, y=mort.rate, color=treatment))
```

Better - but it's best practice to use symbols in addition to colors, in case this
gets printed in black and white, or for colorblind folks. And, our axes come from
the column names - so let's fix those axis labels. And a different theme, or style:

```{r}
ggplot(data=data) +
  geom_point(aes(x=strain, y=mort.rate, color=treatment, shape = treatment)) +
  ylab("Mortality Rate in Bottles") +
  xlab("Mosquito Strain") +
  theme_bw()
```

Nice, clean, and readable.

### Hypothesis Testing
It looks like we might have some differences in mortality rates. Who knew.
We are ready for our hypothesis test. We can use a binomial regression for this.

Our negative control bottles won't go into this, so we can filter them out. They
were mostly there in order to make sure our assays went well, and that we didn't
need to make a correction.

```{r}
glm(mort.rate ~ strain, #Formula, as response.variable ~ explanatory variable.
  weights = total.mosquitoes, #The total number of mosquitoes in each bottle
  data=data %>% filter(treatment=="treatment"), #dataset without NC bottles
  family=binomial) #binomial regression, since these are mortality data.
```

If this is all we do, the function gives output for the fitted coefficients and intercept. Let's create an object, to make some other stuff easier.
```{r}
model.fit <- glm(mort.rate ~ strain,
    weights = total.mosquitoes,
    data=data %>% filter(treatment=="treatment"),
    family=binomial)
```

We can make some handy plots to see how well our model fit.  Yours will never look
this nice, be forwarned!  We simulated this data, so it's pretty synthetic, and if
these didn't look nice I'd really worry that we chose the wrong type of model.
Enter this, then hit enter to keep seeing additional plots. I usually focus on the Q-Q plot and the residual plot.
```{r}
plot(model.fit)
```

This will give you a summary of your model, and some p-values - these can use
different tests by default, depending on the type of model. See ?summary.glm()
```{r}
summary(model.fit)
```

Great - we have data (points), and then fit a line through those points (our model).
These coefficients and intercept describe that line.
```{r}
model.fit
```
Why do we only have coefficients for two of our strains? The short answer is,
the model uses one factor as "reference" factor, fixing it at 0. Check the signs -
the positive coefficient for "tres" means that tres has higher mortality than the reference level (dos, chosen automatically). Coefficient "uno" has a negative sign, meaning that it has lower mortality than dos.  We can think of the intercept as the reference strain here (dos).

### Model Checking
So with these values, we get some predictive power. We can use this handy function
to back-calculate the predicted means for each strain, using the intercept and coefficients from our model:
```{r}
emmeans(model.fit, ~ ".",
        type="response",
        weights = data$total.mosquitoes,
        adjust=TRUE)
```

We simulated these data with each strain having a different mortality rate.
Those were uno-0.25, dos-0.50, tres-0.75.  Compare those with the output of emmeans
(prob column).  Look at the lower and upper confidence limits to see what our
margin of error is. Not bad for just 10 treatment bottles in each group. We can treat non-overlap of these confidence intervals as our hypothesis test.

When your experimental design gets a bit more complex, multiple doses, observation
periods, etc. - we can use predict() to get estimates of mortality from specific
sets of conditions. I'll save that for when we do LC50's.