---
title: "Handy Transformations"
author: "Dan Hartman"
date: "2025-04-07"
output: rmarkdown::github_document
output_file: "README"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300, dev = "svg")
```


## Objectives
1.  Learn a few handy transformations with R functions
2.  Do some more advanced plotting with ggplot2

The last example dataset was pretty simple, and we were able to make plots right
out of the gate. Most folks have also started reading their data into R and doing
some visualization and analysis. Your data is likely going to be a bit more complex
going forward, with multiple observation times, multiple insecticides, etc. This
adds a little complexity to our exploratory analysis. There are many more ways
we can visualize the data - and that's a good thing.

As we go through some of the exploratory analyses, we have to reshape data, add
columns calculated from existing columns, filter data, etc. etc. We've done a bit
of that already. I wanted to give you some recipes for transformations that I use
a lot with bioassay data.

This usually arises due to the fact that data entry is easiest in a wider spreadsheet,
while analysis requires a longer spreadsheet. I'll show below why this is the case and
how we can use some transformation in our R script to automatically do what we
might usually use a pivot table for. This keeps things easier for us both in the
lab and in the R window.

## The Workflow

I expanded the data simulation process from our second meeting to add mortality
at 48 and 72 hour time points - see datasim.R in the meeting folder. These data 
are formatted to match the typical data collection format - see the CDC or WHO 
bioassay manuals as a reference.

After reading in the data, we are going to change the structure of the dataset to be more R-friendly. Going forward, this set of functions should work perfectly to get your data into
a more R friendly format. Just use this example dataset as a template and keep
the column names.

Finally, we will leverage the new format to make a really nice plot using the ggplot2 package in just a few lines of code. Play around with some of the arguments to see what happens.

## Excercise

So, we did all those bioassays, and we want to make some plots. Let's read in
the data. Don't forget to put your data in a project folder and set the working
directory. Go ahead and install ggbeeswarm if you don't already have it.

```{r}
library(tidyverse)
library(ggbeeswarm)
setwd("~/bioassay_analysis_workshop/Meeting_3")
data.wide <- read_csv("three.strain.three.timepoints.csv")
head(data.wide)
```

Looks good - all of the numeric columns are numeric or `dbl`, and the character columns are characters or `chr`.  Great start.

We might be interested in plotting mortality across a a set of time points - but we don't
have that as a variable yet. As a rule of thumb - the best dataset has one
observation per row. Here, we have three observations per row (three time points).
This is going to feel goofy, so let's just do it and then compare the two datasets:

```{r}
data.long<-pivot_longer(data=data.wide,
                        cols = c("dead.hr.24", "dead.hr.48", "dead.hr.72"), #the colums we want to stack
                        names_to = "Time Point", #turn column names into data points
                        values_to = "dead") #move data points to 'dead' column
head(data.long)
```

Now we have `Time Point` as a variable, and our data sheet has one observation per
row.  This would be a pretty ugly table format to record raw data into, in the
insectary. Take a good few minutes to examine `data.long` and `data.wide` to see 
exactly how we transformed the dataset.

Now we can calculate the percent mortality just as before, with `mutate()`. Remember,
`mutate()` adds columns to our table, which can be calculated from existing columns.

```{r}
data.long<-mutate(.data=data.long,
                  mortality = dead / total.mosquitoes) #specify new column and calc from existing ones
head(data.long)
```

We have our mortality calculated - great. I don't like how `Time Point` still has
some characters - lets get rid of that, in a similar way with `mutate()`. This
time, we will overwrite the existing `Time Point` column. We are using `gsub()`
to replace `dead.hr.` with... nothing, in order to remove those existing characters.

```{r}
data.long<-mutate(.data=data.long,
                  'Time Point' = gsub(pattern='dead.hr.', replace='', x=data.long$'Time Point'))
head(data.long)
```

Awesome - it got ride of the words. But see how it is still calling `Time Point` a
character vector?  See the `<chr>` below the column header. Lets fix that.

```{r}
data.long$'Time Point'<-as.numeric(data.long$'Time Point')
head(data.long)
```

Great!  No we have an R friendly data set. Let's do our due diligence and make sure
that control mortality is less than 5%, per the <a href="https://www.who.int/publications/i/item/9789240043770">WHO manual</a>. Above
5%, we need to make some corrections.  Let's see.

```{r}
data.long %>%
  filter(treatment=="negative.control", #remove treatment bottles
         `Time Point`==72) %>% #select only the final time point
  group_by(strain) %>% #specify the groups we want to separate, if any
  summarize(mortality = ( (sum(dead)) / sum(total.mosquitoes) ) ) #summarize mortality across all control bottles
```
Tres comes close to our threshold for needing a correction (5% mortality) but we can proceed without it.

Let's do some plots! Note that with ggplot, we will plot multiple layers, explained below. Anything we put as an argument within `ggplot()` gets used by all "geoms" (layers) in the plot. We want both layers to have the same x and y axes, and be colored by strain, so I just put that in ggplot(). We will add some tweaks to the layers within those specific lines.

If we plot the raw mortality points, we get a lot of overlapping points, so
that ends up being misleading. The `geom_beeswarm` package allows for the points to spread a little bit along the x axis, but only if they overlap. The `cex = 4` argument tells the function how much distance we want - play around with this value and see what happens for you.

`geom_violin()` adds another layer with a symmetrical density describing the data; this is a great way to summarize. The `fill=NA` argument keeps the center transparent so that we can see the raw data points from `geom_beeswarm()`. Try to replace that with `fill=strain` and see what happens.

Finally, `fact_grid()` splits up our data across multiple 'facets' (in this case, according to strain). We describe what we want as a formula, very similar to how we specify regressions. In this case `. ~ strain` says to separate everything (`.`) only by `strain`.
```{r out.width="100%"}
data.long %>%
  filter(treatment=="treatment") %>% #filter out the negative control bottles
ggplot(aes(x=`Time Point`, y=mortality, color=strain, group=`Time Point`)) + #define the axes we want and how we want to group the data
  geom_beeswarm(cex = 4) +
  geom_violin(fill=NA) +
  facet_grid(. ~ strain)
```
See how we could not have dones this without `pivot_longer()`? How would we have specified our time points as a variable in `ggplot()`? If we wanted to include these time points as a variable in a regression, it would be the same situation.

